Comparing fluorometric methods ( in vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds

### 2024-07-19 manuscript/manuscript.qmd
edits based on feedback from MC (BC review)

---
title: "Comparing fluorometric methods ( in vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
author: 
  - name: Stephen D. Shivers
    orcid: 0000-0003-0096-7433
    email: shivers.stephen@epa.gov
    corresponding: true
    affiliation:
      - id: epa
        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jeffrey W. Hollister
    orcid: 0000-0002-9254-9740
    affiliation:
      - ref: epa
  - name: Sophie Fournier
    affiliation:
      - id: orise
        name: Oak Ridge Institute for Science and Education
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jakob Stankoski
    affiliation:
      - id: orau
        name: Oak Ridge Associated Universities
        city: Narragansett
        state: RI
        zip: 02882
  - name: Betty J. Kreakie
    orcid: 0000-0002-4982-7610
    affiliation:
      - ref: epa
abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.  A variety of instruments have been developed to measure cyanoHAB indicators (i.e. chlorophyll *a* or phycocyanin) that do not require advanced laboratory processes (e.g. pigment extraction).  We compared measurements from five in vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.  We found a strong relationship between extracted phycocyanin and  in vivo fluorometers (R^2^ ranging from 0.78-0.96).  We found less consistency between  in vivo measurements of chlorophyll *a* and the extracted results (R^2^ between 0.34 and 0.82).  Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.  Phycocyanin results from  in vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.  Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.  Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.  The results from this study encourage further research using in vivo fluorometry and phycocyanin for cyanoHAB monitoring efforts and when making decisions regarding public health."
format: 
    docx:
      reference-doc: manuscript_template.docx
bibliography: manuscript.bib
editor_options: 
  chunk_output_type: console
---

@@

A variety of instruments have been designed to measure chlorophyll *a* and/or phycocyanin using in vivo fluorescence without an extraction step. These instruments distinguish between the pigments based on their differing emission spectra. Some instruments (i.e. PhycoProbe) use multiple leds at varying wavelengths to cause excitation and emission of different algal type pigments, thus creating a finer scale breakdown of community composition (@beutler_fluorometric_2002). These instruments can be used in the field, the lab, a combination of both, and have a wide range in cost.

@@

![Field sampling locations](images/fig1_map.jpg){#fig-map width="6in" height="8in"}

@@

Within 24 hours of collection, three replicates (i.e. lab replicates to assess precision) from each field sample were analyzed using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction. Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in vivo module. Chlorophyll *a* was extracted by placing frozen GF/F filters in centrifuge tubes with 90% acetone, then placing the tubes for a 20 minute period in a sonicating water bath to lyse cyanobacterial cells (\@welschmeyer1994fluorometric). Phycocyanin was extracted from frozen GF/F filters using a 50 mM phosphate buffer as the extraction solvent and following a 15 minute period in a sonicating water bath to lyse cyanobacterial cells (\@kasinak2015benchtop). Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin).

@@

![Compare Cells/mL of divisions by waterbody](images/fig4_total_cellsml_bar.jpg){#fig-cellsml width="8.5in" height="5.25in"}

@@

The results from this study indicate that in vivo phycocyanin is better related to extracted phycocyanin when compared to the same measurements for chlorophyll. Similarly, phycocyanin from all instruments was better correlated to cell counts than chlorophyll measurements. Instruments measuring in-vivo phycocyanin were better able to handle the cyanobacteria-dominated phytoplankton communities in these study ponds that cause in-vivo chlorophyll measurements to be less strongly correlated with extracted results. Both extracted phycocyanin and chlorophyll were strongly correlated with cell counts across the six study ponds suggesting that pigments are indeed a good proxy for cell counts at least for some waterbodies.

The relationship between phycocyanin and all fluorometers was strong. The Fluorosense had the strongest correlation between the waterbodies, however the Fluorosense overestimated the actual concentrations as compared to extracted values (0-35 ug/L vs 0-12.5 ug/L). Hambdhani et al. also found that the Fluorosense performed well but overestimated chlorophyll *a* at higher algal concentrations (@hamdhani_performance_2021). Therefore, using this particular instrument to enforce set criteria across a variety of waterbodies would probably not be recommended without further research.

The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations compared to extracted concentrations. However, the output from these instruments is not a direct measure of phycocyanin. Instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria, which makes a comparision with other instruments that directly measure phycocyanin difficult. The Cyanofluor output was in RFUs and was not directly comparable to extracted phycocyanin without creating a standard curve for the instrument using known phycocyanin standard concentrations, which was beyond the scope of this study. A study by Thomson-Laing et al. did convert RFUs to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures (@thomson-laing_predicting_2020). They also found stronger relationships for environmental samples in lakes that were cyanobacteria dominated compared to more diverse lakes.

All of these instruments could be used to effectively monitor for the presence of cyanobacteria by measuring more direct parameters of cyanobacteria than chlorophyll *a*. All of the fluorometers exhibited a positive relationship with chlorophyll, but the Algaetorch and Phycoprobe had a much stronger relationship. Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran. Phycocyanin concentrations and cell counts of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (@zamyadi_monitoring_2012). Therefore, the Algaetorch and Phycoprobe are more successful in properly estimating chlorophyll across a range of different waterbodies, but they are also the most expensive instruments. A study by Silva et al. found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (@silva_comparison_2016).

This study supports the use of extracted chlorophyll and phycocyanin as an alternative to cell counts for the study waterbodies. Currently, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments. Therefore, measuring in vivo phycocyanin may be a good proxy for cell counts. There are a variety of instruments capable of measuring phycocyanin using in vivo fluorometry across a spectrum of prices. We hope this study will encourage further research and consideration of using in vivo phycocyanin as alternative to cell counts and in situations were extractive techniques are not able to be used.

@@

We are grateful to Hilary Snook, Darryl Keith, Steve Rego, ACESD QAM, ACESD Supervisor, and Tim Gleason for constructive reviews of this paper. The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S. Environmental Protection Agency. Any mention of trade names, products, or services does not imply an endorsement by the U.S. Government or the U.S. Environmental Protection Agency. The EPA does not endorse any commercial products, services, or enterprises. This contribution is identified by the tracking number ORD-XXXXXX of the Atlantic Coastal Environmental Sciences Division, Office of Research and Development, Center for Environmental Measurement and Modeling, US Environmental Protection Agency. Lastly, special thanks to Gracie, Anastasia, and Lilac for cheer leading and support during challenging analysis times.

>>>>

---
title: "Comparing fluorometric methods ( in vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
author: 
  - name: Stephen D. Shivers
    orcid: 0000-0003-0096-7433
    email: shivers.stephen@epa.gov
    corresponding: true
    affiliation:
      - id: epa
        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jeffrey W. Hollister
    orcid: 0000-0002-9254-9740
    affiliation:
      - ref: epa
  - name: Sophie Fournier
    affiliation:
      - id: orise
        name: Oak Ridge Institute for Science and Education
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jakob Stankoski
    affiliation:
      - id: orau
        name: Oak Ridge Associated Universities
        city: Narragansett
        state: RI
        zip: 02882
  - name: Betty J. Kreakie
    orcid: 0000-0002-4982-7610
    affiliation:
      - ref: epa
abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.  A variety of instruments have been developed to measure cyanoHAB indicators (i.e., chlorophyll *a* or phycocyanin) that do not require advanced laboratory processes (e.g., pigment extraction).  We compared measurements from five in vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.  We found a strong relationship between extracted phycocyanin and  in vivo fluorometers (R^2^ ranging from 0.78-0.96).  We found less consistency between  in vivo measurements of chlorophyll *a* and the extracted results (R^2^ between 0.34 and 0.82).  Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.  Phycocyanin results from  in vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.  Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.  Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.  The results from this study encourage further research using in vivo fluorometry and phycocyanin for cyanoHAB monitoring efforts and when making decisions regarding public health."
format: 
    docx:
      reference-doc: manuscript_template.docx
bibliography: manuscript.bib
editor_options: 
  chunk_output_type: console
---

@@

A variety of instruments have been designed to measure chlorophyll *a* and/or phycocyanin using in vivo fluorescence without an extraction step. These instruments distinguish between the pigments based on their differing emission spectra. Some instruments (i.e., PhycoProbe) use multiple leds at varying wavelengths to cause excitation and emission of different algal type pigments, thus creating a finer scale breakdown of community composition (@beutler_fluorometric_2002). These instruments can be used in the field, the lab, a combination of both, and have a wide range in cost.

@@

![Six field sampling locations in Rhode Island, USA](images/fig1_map.jpg){#fig-map width="6in" height="8in"}

@@

Within 24 hours of collection, three replicates (i.e., lab replicates to assess precision) from each field sample were analyzed using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction. Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in vivo module. Chlorophyll *a* was extracted by placing frozen GF/F filters in centrifuge tubes with 90% acetone, then placing the tubes for a 20 minute period in a sonicating water bath to lyse cyanobacterial cells (\@welschmeyer1994fluorometric). Phycocyanin was extracted from frozen GF/F filters using a 50 mM phosphate buffer as the extraction solvent and following a 15 minute period in a sonicating water bath to lyse cyanobacterial cells (\@kasinak2015benchtop). Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin).

@@

![Cell counts (cells/mL) of phytoplankton samples from six waterbodies in Rhode Island](images/fig4_total_cellsml_bar.jpg){#fig-cellsml width="8.5in" height="5.25in"}

@@

The results from this study indicate that in vivo phycocyanin is better related to extracted phycocyanin when compared to the same measurements for chlorophyll. Similarly, phycocyanin from all instruments was better correlated to cell counts than correlations between chlorophyll measurements from instruments and cell counts. Instruments measuring in-vivo phycocyanin were better able to handle the cyanobacteria-dominated phytoplankton communities in these study ponds that cause in-vivo chlorophyll measurements to be less strongly correlated with extracted results. Both extracted phycocyanin and chlorophyll were strongly correlated with cell counts across the six study ponds, suggesting that pigments are indeed a good proxy for cell counts, at least for some waterbodies.

The relationship between phycocyanin and all fluorometers was strong. The Fluorosense had the strongest correlation between the waterbodies; however, the Fluorosense overestimated the actual concentrations as compared to extracted values (0-35 ug/L vs 0-12.5 ug/L). Hambdhani et al. (2021) also found that the Fluorosense performed well but overestimated chlorophyll *a* at higher algal concentrations. Therefore, using this particular instrument to enforce set criteria across a variety of waterbodies would probably not be recommended without further research.

The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations compared to extracted concentrations; however, the output from these instruments is not a direct measure of phycocyanin. Instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria, which makes a comparision with other instruments that directly measure phycocyanin difficult. The Cyanofluor output was in RFUs and was not directly comparable to extracted phycocyanin without creating a standard curve for the instrument using known phycocyanin standard concentrations, which was beyond the scope of this study. A study by Thomson-Laing et al. (2020) did convert RFUs to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures. They also found stronger relationships for environmental samples in lakes that were cyanobacteria dominated compared to more diverse lakes.

All of these instruments could be used to effectively monitor for the presence of cyanobacteria by measuring more direct parameters of cyanobacteria than chlorophyll *a*. All of the fluorometers exhibited a positive relationship with chlorophyll, but the Algaetorch and Phycoprobe had a much stronger relationship. Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran. Phycocyanin concentrations and cell counts of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were causing underestimation of the in vivo concentrations of chlorophyll (@zamyadi_monitoring_2012). Therefore, the Algaetorch and Phycoprobe are more successful in properly estimating chlorophyll across a range of different waterbodies, but they are also the most expensive instruments. A study by Silva et al. (2016) found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L.

This study supports the use of extracted chlorophyll and phycocyanin as an alternative to cell counts for the study of cyanoHABS. In lieu of extractive techniques, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found in vivo phycocyanin to be better correlated with extracted phycocyanin than correlations between in vivo and extracted chlorophyll a across a variety of instruments. Therefore, measuring in vivo phycocyanin may be a good proxy for cell counts. There are a variety of instruments capable of measuring phycocyanin using in vivo fluorometry across a spectrum of prices. We hope this study will encourage further research and consideration of using in vivo phycocyanin as alternative to cell counts and in situations were extractive techniques are not able to be used.

@@

We are grateful to Hilary Snook, Darryl Keith, Steve Rego, Kara Godineaux, Marty Chintala, and Tim Gleason for constructive reviews of this paper. The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S. Environmental Protection Agency. Any mention of trade names, products, or services does not imply an endorsement by the U.S. Government or the U.S. Environmental Protection Agency. The EPA does not endorse any commercial products, services, or enterprises. This contribution is identified by the tracking number ORD-XXXXXX of the Atlantic Coastal Environmental Sciences Division, Office of Research and Development, Center for Environmental Measurement and Modeling, US Environmental Protection Agency. Lastly, special thanks to Gracie, Anastasia, and Lilac for cheer leading and support during challenging analysis times.
